---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file and knit again -->


```{r, echo = FALSE}
knitr::opts_chunk$set(
  warning = FALSE,   # avoid warnings and messages in the output
  message = FALSE,
  collapse = TRUE,
  fig.width = 5,
  fig.height = 5,
  comment = "#>",
  fig.path = "README-"
)

par(mar=c(3,3,1,1)+.1)
```

```{r, echo=FALSE}
library(mvinfluence)
```

[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/mvinfluence)](https://cran.r-project.org/package=mvinfluence)
[![](http://cranlogs.r-pkg.org/badges/grand-total/mvinfluence)](https://cran.r-project.org/package=mvinfluence)
[![Rdoc](http://www.rdocumentation.org/badges/version/mvinfluence)](http://www.rdocumentation.org/packages/mvinfluence)



# mvinfluence
**Influence Measures and Diagnostic Plots for Multivariate Linear Models**

Functions in this package compute regression deletion diagnostics for multivariate linear models and provides some associated
diagnostic plots.  The diagnostic measures include hat-values (leverages), generalized Cook's distance, and
generalized squared 'studentized' residuals.  Several types of plots to detect influential observations are
provided.

## Installation

Get the released version from CRAN:

     install.packages("mvinfluence")

## Goals

The design goal for this package is that, as an extension of standard methods for univariate linear models, you should
be able to fit a linear model with a multivariate response,

    mymlm <- lm( cbind(y1, y2, y3) ~ x1 + x2 + x3, data=mydata)

and then get useful diagnostics and plots with:

    influence(mymlm)
    hatvalues(mymlm)
    influencePlot(mymlm, ...)  

## Examples

Fit a MLM to a subset of the Rohwer data (the Low SES group).  
The default influence plot (`type="stres"`) shows the squared standardized residual against the Hat value.
The areas of the circles representing the observations are proportional to generalized Cook's distances. 
```{r rohwer1}
data(Rohwer, package="heplots")
Rohwer2 <- subset(Rohwer, subset=group==2)
Rohwer.mod <- lm(cbind(SAT, PPVT, Raven) ~ n+s+ns+na+ss, data=Rohwer2)

influencePlot(Rohwer.mod, id.n=4)
```

An alternative (`type="stres"`) plots residual components against leverage components,
with the property that contours of constant Cook's distance fall on diagonal
lines with slope = -1.

```{r rohwer2}
influencePlot(Rohwer.mod, id.n=4, type="LR")
```




